Miniproject - Algebraic Geometry

### 2021-12-12 Algebraic geometry/submission.tex
Completely and rigorously reproved the independence lemma

\section{Part I}

\begin{definition}
    Let $V$ be a vector space.
    Then define the $d$-th exterior power as
    \begin{equation*}
        \extpow^d(V) := V^{\otimes d} \ / \ \sum_{i = 1}^{d - 1} V^{\otimes (i - 1)} \otimes \vspan\bigl\{ v \otimes v' + v' \otimes v \ \bigm| \ v, v' \in V \bigr\} \otimes V^{\otimes (d - i - 1)}
    \end{equation*}
    Use the notation $v_1 \wedge ... \wedge v_d := [v_1 \otimes ... \otimes v_d] \in \extpow^k(V)$.
\end{definition}

\begin{lemma}
    \label{prop:basic_properties_exterior_product}
    Let $v_1, ..., v_d \in V$. Have for $\pi \in S_d$ that
    \begin{equation*}
        v_{\pi(1)} \wedge ... \wedge v_{\pi(k)} = \sgn(\pi) (v_1 \wedge ... \wedge v_d)
    \end{equation*}
    Furthermore if $v_i = v_j$ for some $i \neq j$, then
    \begin{equation*}
        v_1 \wedge ... \wedge v_d = 0
    \end{equation*}
\end{lemma}
\begin{proof}
    Note that
    \begin{equation*}
        u \wedge v \wedge v' \wedge w = -(u \wedge v' \wedge v \wedge w)
    \end{equation*}
    for all $u \in \extpow^{i - 1}(V), w \in \extpow^{(d - i - 1)}(V), v, v' \in V$.

@@

\begin{lemma}
    \label{prop:linear_transform_extpow}
    Let $A = (a_{ij}) \in \GL_d(k)$ and $v_1, ..., v_d \in V$.
    Then
    \begin{equation*}
        \Bigl( \sum_j a_{1j}v_j \Bigl) \wedge ... \wedge \Bigl( \sum_{j} a_{dj} v_j \Bigr) = \det(A) (v_1 \wedge ... \wedge v_d)
    \end{equation*}
\end{lemma}
\begin{proof}
    By a direct computation using Lemma~\ref{prop:basic_properties_exterior_product}, we find
    \begin{align*}
        &\Bigl( \sum_j a_{ij} v_j \Bigr) \wedge ... \wedge \Bigl( \sum_j a_{dj} v_j \Bigr) = \sum_{j_1, ..., j_d} a_{1j_1} ... a_{dj_d} (v_{j_1} \wedge ... \wedge v_{j_d}) \\
        = &\sum_{\pi \in S_d} a_{1\pi(1)} ... a_{d\pi(d)} (v_{\pi(1)} \wedge ... \wedge v_{\pi(d)}) \\
        = &\sum_{\pi \in S_d} a_{1\pi(1)} ... a_{d\pi(d)} \sgn(\pi) (v_1 \wedge ... \wedge v_d) \\
        = &(v_1 \wedge ... \wedge v_d) \ \sum_{\pi \in S_d} \sgn(\pi) \prod_{j = 1}^d a_{j\pi(j)} = \det(A) (v_1 \wedge ... \wedge v_d)
    \end{align*}
    where the last equality holds due to the Leibniz determinant formula.
\end{proof}
\begin{lemma}
    \label{prop:extpow_zero_iff_independent}
    For $v_1, ..., v_d \in V$ have
    \begin{equation*}
        v_1 \wedge ... \wedge v_d = 0 \ \Leftrightarrow \ v_1, ..., v_d \ \text{linearly dependent}
    \end{equation*}
\end{lemma}
\begin{proof}
    For the direction $\Leftarrow$, assume that $v_1, ..., v_d$ are not independent. 
    Then there is a nonzero vector $a_1 \in k^d$ with $\sum a_{1i} v_i = 0$.
    Clearly, we can extend $a_1$ to a basis $a_1, ..., a_d$ of $k^d$, which gives a matrix $A = (a_{ij}) \in \GL_d(k)$.

@@

    Direction $\Rightarrow$ TODO
\end{proof}
\begin{lemma}
    \label{prop:subspace_closed}
    Let $v \in V$ and $u \in \extpow^{d - 1} U$ for a linear subspace $U \leq V$. 
    If $v \wedge u \in \extpow^d U$ then $v \in U$ or $u = 0$.
\end{lemma}
\begin{proof}
    TODO
\end{proof}

>>>>

For this document, let $V$ be a finitely dimensional vector space with basis $B = (b_1, ..., b_n)$.

\section{Part I}
First of all, I will include here a more formal definition of the exterior product, which is required for rigorous proofs later on.
\begin{definition}
    Define the $d$-th exterior power of $V$ as a quotient of the free vector space of $B \times ... \times B$ as follows
    \begin{align*}
        \extpow^d V := \mathrm{Fr}\Bigl(\bigtimes_{i = 1}^d B\Bigr) \ \bigm/ \ U
    \end{align*}
    where
    \begin{align*}
        U = \vspan \bigl\{ (v_1, ..., v_d) + (v_1, ..., v_{i - 1}, v_{i + 1}, v_i, v_{i + 2}, ..., v_d) \ \bigm| \ v_1, ..., v_d \in B, 1 \leq i < d \bigr\}
    \end{align*}
    Consider also the map
    \begin{equation*}
        \wedge: \bigtimes_{i = 1}^d V \to \extpow^d V, \quad \Bigl( \sum_{i = 1}^n \lambda_{1i} b_i, ..., \sum_{i = 1}^n \lambda_{di} b_i \Bigr) \mapsto \sum_{1 \leq i_1, ..., i_d \leq n} \lambda_{1i_1} ... \lambda_{di_d} (b_{i_1}, ..., b_{i_d})
    \end{equation*}
    and use the notation $v_1 \wedge ... \wedge v_d$ to denote the image of $(v_1, ..., v_d)$ under this map.
\end{definition}
First of all, we show some basic properties of the exterior product.
It is straightforward to see that the exterior product is multilinear, i.e. $v_1 \wedge ... \wedge (v_i + v'_i) \wedge ... \wedge v_d = (v_1 \wedge ... \wedge v_d) + (v_1 \wedge ... \wedge v'_i \wedge ... \wedge v_d)$.
More interesting are the following properties.
\begin{lemma}
    \label{prop:basic_properties_exterior_product}
    Let $v_1, ..., v_d \in V$. Have for $\pi \in S_d$ that
    \begin{equation*}
        v_{\pi(1)} \wedge ... \wedge v_{\pi(k)} = \sgn(\pi) (v_1 \wedge ... \wedge v_d)
    \end{equation*}
    Furthermore if $v_i = v_j$ for some $i \neq j$, then
    \begin{equation*}
        v_1 \wedge ... \wedge v_d = 0
    \end{equation*}
\end{lemma}
\begin{proof}
    Clearly the vectors $b_{i_1} \wedge ... \wedge b_{i_d}$ span $\extpow^d V$.
    So let
    \begin{equation*}
        u = \sum_{i_1, ..., i_{j - 1}} \lambda_{i_1, ..., i_{j - 1}} (b_1 \wedge ... \wedge b_{j - 1}), \quad w = \sum_{i_1, ..., i_{d - j - 1}} \mu_{i_1, ..., i_d} (b_{i_1} \wedge ... \wedge b_{d - j - 1})
    \end{equation*}
    be arbitrary vectors in $\extpow^{j - 1} V$ resp. $\extpow^{d - j - 1} V$ for some $1 \leq j \leq d - 1$.
    Note that for $v = \sum \tau_i b_i, v' = \sum \tau'_i b_i \in V$ have then
    \begin{align*}
        &(u \wedge v \wedge v' \wedge w) + (u \wedge v' \wedge v \wedge w) = \sum_{i_1, ..., i_{j - 1}, i_{j + 2}, ..., i_d} \quad \sum_{0 \leq i, i' \leq n} \lambda_{i_1, ..., i_{j - 1}} \mu_{i_{j + 2}, ..., i_d} \tau_i \tau'_{i'} \\
        &\Bigl(\underbrace{\bigl( b_{i_1} \wedge ... \wedge b_{i_j} \wedge b_{i_{j + 1}} \wedge ... \wedge b_{i_d}\bigr) + \bigl( b_{i_1} \wedge ... \wedge b_{i_{j - 1}} \wedge b_{i_{j + 1}} \wedge b_{i_j} \wedge b_{i_{j + 2}} \wedge ... \wedge b_{i_d} \bigr)}_{= 0 \ \text{as it is an element of the space $U$}} \Bigl) \\
        &= 0
    \end{align*}
    for all $u \in \extpow^{i - 1}(V), w \in \extpow^{(d - i - 1)}(V)$.
    Hence
    \begin{equation*}
        u \wedge v \wedge v' \wedge w = -(u \wedge v' \wedge v \wedge w)
    \end{equation*}

@@

\begin{lemma}
    \label{prop:linear_transform_extpow}
    Let $A = (a_{ij}) \in \GL_d(k)$ and $v_1, ..., v_d \in V$.
    Then
    \begin{equation*}
        \Bigl( \sum_j a_{1j}v_j \Bigl) \wedge ... \wedge \Bigl( \sum_{j} a_{dj} v_j \Bigr) = \det(A) (v_1 \wedge ... \wedge v_d)
    \end{equation*}
\end{lemma}
\begin{proof}
    By a direct computation using Lemma~\ref{prop:basic_properties_exterior_product}, we find
    \begin{align*}
        &\Bigl( \sum_j a_{ij} v_j \Bigr) \wedge ... \wedge \Bigl( \sum_j a_{dj} v_j \Bigr) = \sum_{j_1, ..., j_d} a_{1j_1} ... a_{dj_d} (v_{j_1} \wedge ... \wedge v_{j_d}) \\
        = &\sum_{\pi \in S_d} a_{1\pi(1)} ... a_{d\pi(d)} (v_{\pi(1)} \wedge ... \wedge v_{\pi(d)}) \\
        = &\sum_{\pi \in S_d} a_{1\pi(1)} ... a_{d\pi(d)} \sgn(\pi) (v_1 \wedge ... \wedge v_d) \\
        = &(v_1 \wedge ... \wedge v_d) \ \sum_{\pi \in S_d} \sgn(\pi) \prod_{j = 1}^d a_{j\pi(j)} = \det(A) (v_1 \wedge ... \wedge v_d)
    \end{align*}
    where the last equality holds due to the Leibniz determinant formula.
\end{proof}
The next two lemmas are quite fundamental, as they can be used as a basic tool to prove linear independence in the exterior product.
\begin{lemma}
    \label{prop:basis_product_independence}
    The vectors $b_{i_1} \wedge ... \wedge b_{i_d}$ for $i_1 < ... < i_d$ form a basis of $\extpow^d V$.
\end{lemma}
\begin{proof}
    This proof is slightly technical, as I wanted to provide a rigorous proof using only the definition of $\extpow^d V$ and some properties of the symmetric group $S_d$.
    The core idea is to show the linear independence by going to down to the free product underlying $\extpow^d V$, and then show that we cannot use the vectors $(b_{i_1}, ..., b_{i_d}) + (b_{i_1}, ..., b_{i_{j + 1}}, b_{i_j}, ..., i_d)$ to create a nontrivial linear combination.
    Namely, assuming we have such a nontrivial linear combination, we can group it into the part where the order of the $i_1, ..., i_d$ has even/odd parity.
    The vectors with $i_1 < ... < i_d$ clearly are all on one side, and from this we can derive the claim.

    Note that by definition, the vectors $b_{i_1} \wedge ... \wedge b_{i_d}$ for any $1 \leq i_1, ..., i_d \leq n$ span $\extpow^d V$.
    By Lemma~\ref{prop:basic_properties_exterior_product}, we can rewrite any of them (that is nonzero) to be of the above form, so clearly the $b_{i_1} \wedge ... \wedge b_{i_d}$ for $i_1 < ... < i_d$ span $\extpow^d V$.

    It is left to show that they are linearly independent.
    Assume
    \begin{equation*}
        \sum_{i_1 < ... < i_d} \lambda_{i_1, ..., i_d} \bigl( b_{i_1} \wedge ... \wedge b_{i_d} \bigr) = 0
    \end{equation*}
    Then clearly
    \begin{equation*}
        \sum_{i_1 < ... < i_d} \lambda_{i_1, ..., i_d} \bigl( b_{i_1}, ..., b_{i_d} \bigr) \in U
    \end{equation*}
    where $U$ is the vector space from the definition of $\extpow^d V$.
    So
    \begin{align*}
        \sum_{i_1 < ... < i_d} \lambda_{i_1, ..., i_d} \bigl( b_{i_1}, ..., b_{i_d} \bigr) = &\sum_{j, \ i_1, ..., i_d} \mu_{j, i_1, ..., i_d} \\
        &\Bigl( \bigl( b_{i_1}, ..., b_{i_d} \bigr) + \bigl( b_{i_1}, ..., b_{i_{j - 1}}, b_{i_{j + 1}}, b_{i_j}, b_{i_{j + 2}}, ..., b_{i_d} \bigr) \Bigr)
    \end{align*}
    Note that we can assume wlog that the sum on the right-hand side goes only over $i_1, ..., i_d$ distinct.
    The reason is that the other vectors on the right-hand side are contained in
    \begin{equation*}
       \vspan \{ (b_{i_1}, ..., b_{i_d}) \ | \ \text{$i_1, ..., i_d$ not distinct} \}
    \end{equation*}
    which only trivially intersects the space
    \begin{equation*}
        \vspan\{ (b_{i_1}, ..., b_{i_d}) \ | \ \text{$i_1, ..., i_d$ distinct} \}
    \end{equation*}
    due to the properties of the free product (the spaces share now basis vectors).

    Now consider for $i_1, ..., i_d$ distinct the ``sorting permutation'' $\sigma_i \in S_d$, which is the unique permutation such that $i_{\sigma(1)} < ... < i_{\sigma(d)}$.
    Then
        \sum_{i_1 < ... < i_d} \lambda_{i_1, ..., i_d} \underbrace{\bigl( b_{i_1}, ..., b_{i_d} \bigr)}_{\sgn(\sigma_i) = 1 \ \text{as} \ \sigma_i = \mathrm{id}} = &\sum_{\substack{j, \ i_1, ..., i_d\\\sgn(\sigma_i) = 1}} \underbrace{\sgn(\sigma_i) \bigl( \mu_{j, i_1, ..., i_d} - \mu_{j, i_1, ..., i_{j + 1}, i_j, ..., i_d} \bigr)}_{=: \mu'_{j, i_1, ..., i_d}} \\
        &\Bigl( \bigl( b_{i_1}, ..., b_{i_d} \bigr) + \bigl( b_{i_1}, ..., b_{i_{j - 1}}, b_{i_{j + 1}}, b_{i_j}, b_{i_{j + 2}}, ..., b_{i_d} \bigr) \Bigr)
    This yields
        &-\sum_{\substack{j, \ i_1, ..., i_d\\\sgn(\sigma_i) = 1}} \Bigl( \mu'_{j, i_1, ..., i_d} - \begin{cases}
            \lambda_{i_1, ..., i_d} & \text{if $i_1 < ... < i_d, j = 0$} \\
            0 & \text{otherwise}
        \end{cases} \Bigr) (b_{i_1}, ..., b_{i_d}) \\
        =& \sum_{\substack{j, \ i_1, ..., i_d\\\sgn(\sigma_i) = -1}} \mu'_{j, i_1, ..., i_{j + 1}, i_j, ..., i_d} (b_{i_1}, ..., b_{i_d})
    However, all vectors on the left-hand side are contained in
        \vspan \{(b_{i_1}, ..., b_{i_d}) \ | \ \sgn(\sigma_i) = 1 \}
    and all vectors on the right-hand side are contained in
        \vspan \{(b_{i_1}, ..., b_{i_d}) \ | \ \sgn(\sigma_i) = -1 \}
    These two spaces intersect trivially, and so find
        \sum_{\substack{j, \ i_1, ..., i_d\\\sgn(\sigma_i) = -1}} \mu'_{j, i_1, ..., i_{j + 1}, i_j, ..., i_d} (b_{i_1}, ..., b_{i_d}) = 0
    Clearly the different $(b_{i_1}, ..., b_{i_d})$ are linearly independent, so find that all $\mu'_{j, i_1, ..., i_d} = 0$.
    Observe now that thus all $\lambda_{i_1, ..., i_d} = 0$ and the claim is shown.
    \label{prop:subspace_closed}
    Let $v \in V$ and $w \in \extpow^{d - 1} W$ for a linear subspace $W \leq V$. 
    If $v \wedge w \in \extpow^d W$ then $v \in W$ or $w = 0$.
    Here
        \extpow^d W := \{ w_1 \wedge ... \wedge w_d \ | \ w_i \in W \} \leq \extpow^d V
    is a subspace of $\extpow^d V$.
    wlog assume that $b_1, ..., b_m$ are a basis of $U$.
    Then
        w = \sum_{i_1 < ... < i_{d - 1} < m} \lambda_{i_1, ..., i_{d - 1} \leq m} (b_{i_1} \wedge ... \wedge b_{i_{d - 1}})
    So with $v = \sum_i \mu_i b_i$ get
        v \wedge w =& \sum_{i_1 < ... < i_{d - 1} \leq m} \lambda_{i_1, ..., i_{d - 1}} (v \wedge b_{i_1} \wedge ... \wedge b_{i_{d - 1}}) \\
        =& \sum_{i_1 \leq n, \ i_1 < ... < i_{d - 1} \leq m} \mu_{i_1} \lambda_{i_2, ..., i_d} (b_{i_1} \wedge ... \wedge b_{i_d})
    By assumption, we also find
        v \wedge w =& \sum_{i_1 < ... < i_d \leq m} \tau_{i_1, ..., i_d} (b_{i_1} \wedge ... \wedge b_{i_d})
    Both representations are equal, hence
        \sum_{i_1 \leq n, \ i_2 < ... < i_d \leq m} \Bigl( \mu_{i_1}\lambda_{i_2, ..., i_d} - \begin{cases}
            \tau_{i_1, ..., i_d} & \text{if $i_1 \leq m$} \\
            0 & \text{otherwise}
        \end{cases} \Bigr) (b_{i_1} \wedge ... \wedge b_{i_d}) = 0
    Hence by the previous Lemma~\ref{prop:basis_product_independence}, we see that for all $i_2 < ... < i_d \leq m < i_1 \leq n$ have
        0 = \mu_{i_1}\lambda_{i_2, ..., i_d} - \begin{cases}
            \tau_{i_1, ..., i_d} & \text{if $i_1 \leq m$} \\
            0 & \text{otherwise}
        \end{cases}
    So in particular find for all $i_2, ..., i_d \leq m < i_1 \leq n$ that
        \mu_{i_1} \lambda_{i_2, ..., i_d} = 0
    Hence either $\mu_{i_1} = 0$ for all $i_1 > m$ and so $v \in U$, or all $\lambda_{i_2, ..., i_d} = 0$ (for $i_2 < ... < i_d \leq m$) and so $w = 0$.
    \label{prop:extpow_zero_iff_independent}
    For $v_1, ..., v_d \in V$ have
        v_1 \wedge ... \wedge v_d = 0 \ \Leftrightarrow \ v_1, ..., v_d \ \text{linearly dependent}
    For the direction $\Leftarrow$, assume that $v_1, ..., v_d$ are not independent. 
    Then there is a nonzero vector $a_1 \in k^d$ with $\sum a_{1i} v_i = 0$.
    Clearly, we can extend $a_1$ to a basis $a_1, ..., a_d$ of $k^d$, which gives a matrix $A = (a_{ij}) \in \GL_d(k)$.

@@

    For the other direction, let $v_1, ..., v_d$ be linearly independent.
    TODO
\end{proof}
